---
title: "Getting odin up and running on your computer"
author: "Amy Dighe"
date: "27 February 2018"
output: html_document
---
## how to install *odin*

1. Make sure you have a [compiler](https://cran.r-project.org/bin/windows/Rtools/)
    + If you don't, install the one listed for your operating system and make sure you tick to  "add Rtools to 
the PATH"
    + (ask Rich how you could check if you have a compiler without waiting for the error message?)

2. Install the *odin* package as follows in the code chunk below:

```{r, eval = FALSE}
# install.packages("drat") # if the next line fails
drat:::add("mrc-ide")
install.packages("odin")

```

NOTE! if you don't have a compiler you may get the following error message and you will need to return to step 1:
```r
Package which is only available in source form, and may need compilation of C/C++/Fortran: ‘cinterpolate’
  These will not be installed
trying URL 'https://cran.rstudio.com/bin/windows/contrib/3.4/deSolve_1.20.zip'
Content type 'application/zip' length 2667305 bytes (2.5 MB)
downloaded 2.5 MB

```
...

```r
ERROR: dependency 'cinterpolate' is not available for package 'odin'
* removing 'C:/Users/ad3716/Documents/R/win-library/3.4/odin'
Warning in install.packages :
  running command '"C:/PROGRA~1/R/R-34~1.2/bin/x64/R" CMD INSTALL -l "C:\Users\ad3716\Documents\R\win-library\3.4" C:\Users\ad3716\AppData\Local\Temp\RtmpEPC21k/downloaded_packages/odin_0.0.2.tar.gz' had status 1
Warning in install.packages :
  installation of package ‘odin’ had non-zero exit status
```

3. In order to have full use of *odin* you will also need to install the package *dde*

```{r, eval = FALSE}
devtools::install_github("richfitz/dde")
```